Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-07-22 Report/Sections/simulation.tex

Five hundred separate sequences of 100 killer whale dives were simulated according to an HMM, where the hidden Markov chain $X$ was a collection of dive types and the observations $Y$ were the corresponding dive durations (in seconds). Each dive could be one of $N=2$ dive types, and the duration of the $t^{th}$ dive, $Y_t$, followed a gamma distribution whose parameters $\theta^{(i)}$ were dependent on the dive type $X_t = i$. We parameterize the gamma distribution by its mean and variance:
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}, \qquad \delta =  \begin{pmatrix} 0.5 & 0.5 \end{pmatrix},$$
$$Y_t|X_t \sim \rm{Gamma},$$
\begin{align*}
	&\bbE(Y_t|X_t = 1) = 15 s, &\bbE(Y_t|X_t = 2) = 60 s, \\
	%
	&\bbV(Y_t|X_t = 1) = 25 s^2, &\bbV(Y_t|X_t = 2) = 100 s^2.
\end{align*}
%
Once the dive durations were calculated for all 100 dives, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments (the end of the dive sequence was discarded) which made up a second fine-scale hidden Markov model. Each two second segment was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$. The probability transition matrices for these fine-scale Markov chains were set as
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.9 & 0.1 \end{pmatrix}, \qquad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix},$$ 
%
where $\Gamma^{*(1)}$ was used for dives where $X_t = 1$ and $\Gamma^{*(2)}$ was used for dives where $X_t = 2$.

@@

$$\mathbb{E}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \phi_1^{*(i)}z + (1-\phi_1^{*(i)}) \mu_1^{*(i)}$$
$$\mathbb{V}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_1^{*(i)}\right)^2.$$
%
The distribution of $Z^{*(2)}_{t,t^*}$ is gamma and its parameters depend upon only $X^*_{t,t^*}$ (not $X_t$):
%
$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(2)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(2)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%
None of the fine-scale emission distributions depend upon dive type. In total the parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters), and} \\
    %
    \Theta^* = \{\Theta^{*(1)},\Theta^{*(2)}\}  \qquad \text{(fine-scale emission parameters), where} \\
    %
    \Theta^{*(i^*)} =  \{\{\mu_1^{*(i^*)},\sigma_1^{*(i^*)},\phi_1^{*(i^*)}\},\{\mu_2^{*(i^*)},\sigma_2^{*(i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
\end{gather*}
%
The likelihood of this model is still easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model.

Four different models were fit to the simulated data sets:
\begin{enumerate}
    \item A \textbf{CarHHMM-DFT} as described above.
    \item An \textbf{HHMM-DFT}, which is similar to the model above, but with no modeled auto-correlation, i.e. $\phi_1^{*(i^*)} = 0$ for $i^* = 1,2$.
    \item A \textbf{CarHHMM}, which is similar to the model above, but without $Z^{*(2)}$ as an observation (i.e $\Theta^{*(i^*)} = \{\mu_1^{(i^*)},\sigma_1^{(i^*)},\phi_1^{(i^*)}\}$, $i^* = 1,2$).
    \item A \textbf{CarHMM-DFT}, which is similar to the CarHHMM-DFT, but with $N=1$ instead of $N=2$ on the coarse scale. This is equivalent to loosing the coarse level of the hierarchical structure, as there is only one dive type.
\end{enumerate}
%
Each of the last three models leaves out one important aspect of the full CarHHMM-DFT. The CarHMM-DFT lacks a hierarchical structure, the HHMM-DFT is missing auto-correlation within the fine-scale observations, and the CarHHMM does not have access to the Fourier transform sums $(Z^{*(2)})$ as observations.

@@

All models were run on the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
%%% Accuracy %%%
Each model was able to decode the fine-scale hidden states of the process almost perfectly except for the CarHHMM, whose accuracy was 89\%. The lower accuracy of the CarHHMM can be attributed to the higher fine-scale states variation in the distribution of $Z^{*(2)}$ compared to $Z^{*(1)}$. For the coarse-scale hidden states, the CarHMM-DFT lacked a hierarchical structure and could not make any predictions at all. The other three models all achieved an accuracy of approximately 90\%, with the CarHHMM slightly more likely to categorize a dive as dive type 2 than the other models. Figure \ref{fig:acc} shows the decoded state probabilities for both the fine- and coarse- scales for one of the simulation run. Table \ref{table:accuracy} lists the accuracy and training times of each model.

%%% dive duration %%%
For the emission distributions of dive duration, Figure \ref{fig:MLE_dist} shows the empirical joint density of $\hat \mu$ and $\hat \sigma$ for the CarHHMM-DFT. The supplementary material shows the empirical joint densities for all other models and features.

@@

%%% Acceleration %%%
For acceleration ($Z^{*(1)}$), both the CarHMM-DFT and the CarHHMM-DFT regularly converged to the correct parameters with very little standard error. However, the estimated standard error based on the observed Fisher information regularly overestimated the empirical standard error of $\hat \mu$ for both of these models. 
%This likely is due to the very large auto-correlation terms ($\phi^{(1)} = 0.99$ and $\phi^{(2)} = 0.95$) since as $\phi \to 1$, the value of $\mu$ does not matter in the likelihood evaluation as $\phi \to 1$ implies that $\mathbb{E}(Z^{*(1)}_{t,s^*}) = \phi*(Z^{*(1)}_{t,s^*-1}) + (1-\phi)\mu \to Z^{*(1)}_{t,s^*-1}$. 
The HHMM-DFT regularly overestimates the variance $Z^{*(1)}$ since it does not incorporate auto-correlation into the emission distribution of $Z^{*(1)}$. The CarHHMM has large biases in many of its parameter estimates, especially in sub-dive state 2. 
%because it sees the ``wigliness" of the acceleration data as raw acceleration rather than as Fourier modes, which deflates the value of $\hat \phi$ and inflates the value of $\hat \sigma$. 
See the supplementary material for a detailed breakdown of parameter values for acceleration emission distributions.

@@

>>>>

Five hundred separate sequences of 100 killer whale dives were simulated according to an HMM, where the hidden Markov chain $X$ was a collection of dive types and the observations $Y$ were the corresponding dive durations (in seconds). Each dive could be one of $N=2$ dive types, and the duration of the $t^{th}$ dive, $Y_t$, followed a gamma distribution whose parameters $\theta^{(i)}$ were dependent on the dive type $X_t = i$. We parameterize the gamma distribution by its mean and variance:
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}, \qquad \delta =  \begin{pmatrix} 0.5 & 0.5 \end{pmatrix},$$
$$Y_t|X_t \sim \rm{Gamma},$$
\begin{align*}
	&\bbE(Y_t|X_t = 1) = 15 s, &\bbE(Y_t|X_t = 2) = 60 s, \\
	%
	&\bbV(Y_t|X_t = 1) = 25 s^2, &\bbV(Y_t|X_t = 2) = 100 s^2.
\end{align*}
%
Once the dive durations were calculated for all 100 dives, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments, which made up a second fine-scale hidden Markov model. The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive were discarded. Each two-second segment was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$. The probability transition matrices for these fine-scale Markov chains were set as
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix}, \qquad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix},$$ 
%
where $\Gamma^{*(1)}$ was used for dives where $X_t = 1$ and $\Gamma^{*(2)}$ was used for dives where $X_t = 2$.

@@

$$\mathbb{E}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \phi_1^{*(i)}z + (1-\phi_1^{*(i)}) \mu_1^{*(i)}$$
$$\mathbb{V}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_1^{*(i)}\right)^2.$$
%
The distribution of $Z^{*(2)}_{t,t^*}$ is gamma and its parameters depend upon only $X^*_{t,t^*}$ (not $Z^{*(2)}_{t,t^*-1}$):
%
$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(2)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(2)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%
None of the fine-scale emission distributions depend upon dive type. In total the parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters), and} \\
    %
    \Theta^* = \{\Theta^{*(1)},\Theta^{*(2)}\}  \qquad \text{(fine-scale emission parameters), where} \\
    %
    \Theta^{*(i)} =  \{\{\mu_1^{*(i)},\sigma_1^{*(i)},\phi_1^{*(i)}\},\{\mu_2^{*(i)},\sigma_2^{*(i)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
\end{gather*}
%
The likelihood of this model is still easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model.

Four different models were fit to the simulated data sets:
\begin{enumerate}
    \item A \textbf{CarHHMM-DFT} as described above.
    \item An \textbf{HHMM-DFT}, which is similar to the CarHHMM-DFT, but with no modeled auto-correlation \big(i.e. $\phi_1^{*(i)} = 0$ for $i = 1,2$\big).
    \item A \textbf{CarHHMM}, which is also similar to the CarHHMM-DFT, but without the Fourier sums $(Z^{*(2)})$ as an observation \big(i.e $\Theta^{*(i)} = \{\mu_1^{(i)},\sigma_1^{(i)},\phi_1^{(i)}\}$, $i = 1,2$\big).
    \item A \textbf{CarHMM-DFT}, which is again similar to the CarHHMM-DFT, but with $N=1$ instead of $N=2$ on the coarse scale. This is equivalent to loosing the coarse level of the hierarchical structure, as there is only one dive type.
\end{enumerate}
%
Each of the last three models leaves out one important aspect of the full CarHHMM-DFT. The CarHMM-DFT lacks a hierarchical structure, the HHMM-DFT is missing auto-correlation within the fine-scale observations, and the CarHHMM does not have access to the Fourier transform sums $(Z^{*(2)})$ as observations.

All models were run on the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.

@@

%%%

As expected, the full CarHHMM-DFT was the best-performing model. It was able to decode the fine-scale hidden states of the process almost perfectly (mean error $< 10^{-12}$ for all simulations) and decode the hidden dive types with an average accuracy of $94 \pm 3 \%$, where $\pm$ refers to the standard deviation across simulations (see Table \ref{table:accuracy}). All parameter estimates have biases which are at least approximately one order of magnitude smaller than the associated standard error, which is consistent with a sample size of 500 (section S-1.2, tables 2-5). One exception is for estimates of $\hat \sigma$, which were regularly underestimated. For dive duration $(Y_t)$ this finding is consistent with properties of MLEs for standard deviation, especially with a small sample size. For acceleration $(Z^{*(1)})$ this is likely due to over-fitting the auto-correlation parameter $\phi$. The Fourier sums $(Z^{*(2)})$ show no significant bias in $\hat \sigma^*_2$. The empirical standard error of all parameters estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) were well-approximated by the inverse of the observed fisher information matrix (section S-1.2, tables 2-5). One exception is $\hat \Theta$, whose standard errors were slightly underestimated using the observed fisher information matrix. 
%The empirical standard error of $\hat \mu_2$ was 8.5, while the estimate of standard error using the observed Fisher information was 5.8. 
See Section S-1.2, tables 2-5 for more information.

The HHMM-DFT performed almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias. 
%- it had very low error for the fine-scale hidden states and $\approx 94\%$ accuracy on the dive types (see Table \ref{table:accuracy}). In addition, the standard errors of the parameter estimates were comparable to those of the CarHHMM-DFT. The estimates of standard error using the observed Fisher information were also comparable to the CarHHMM-DFT. 
The HHMM-DFT \textit{did} differ from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$. For both sub-dive states, $\hat \sigma_1^*$ greatly overestimated the true standard deviation. In particular, $\bar \hat \sigma_1^{*(1)} = 0.248$ and $\bar \hat \sigma_1^{*(2)} = 0.242$, but $\sigma_1^{*(1)} = 0.05$ and $\sigma_1^{*(2)} = 0.1$. in addition, the empirical standard error of both $\hat \mu_1^*$ and $\hat \sigma_1^*$ was much larger than the estimates of standard error using the observed fisher information (see Section S-1.2, Table 3). This implies that parameter uncertainty can be underestimated when auto-correlation is ignored.

The CarHHMM was the worst-performing model in terms of decoding accuracy, with an accuracy of $89 \pm 2\%$ when determining sub-dive behavior and an accuracy of $91 \pm 3\%$ when determining dive types. It was much more likely to misclassify sub-dive behavior 1 than sub-dive behavior 2 (see Section S-1.2, Table 1). It was also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy. $\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ in particular were poor estimates of the true parameters (see Section S-1.1.3).

Finally, the CarHMM-DFT was comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations. In addition, the time required to fit the CarHHMM was approximately 1/5 that of the other models. However, it lacked any hierarchical structure, so it was unable to decode the coarse-scale dive types. The CarHMM-DFT still did fit a gamma distribution to the dive duration of \textit{all} dives. The parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ is a significant underestimate by $\approx$ 5 seconds.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% EVERYTHING BELOW HERE IS OLD AND I THINK IT IS BAD. THE STUFF ABOVE IS MUCH BETTER.

\iffalse
%%% Accuracy %%%
Each model was able to decode the fine-scale hidden states of the process almost perfectly except for the CarHHMM, whose accuracy was 89\% (Table \ref{table:accuracy}). The lower accuracy of the CarHHMM (i.e. model without $Z^{*(2)}$)  can be attributed to the higher fine-scale states variation in the distribution of $Z^{*(2)}$ compared to $Z^{*(1)}$. For the coarse-scale hidden states, the CarHMM-DFT lacked a hierarchical structure and could not make coarse-scale predictions. All three other models achieved an accuracy of approximately 90\% (Table \ref{table:accuracy}), with the CarHHMM slightly more likely to categorize a dive as dive type 2 than the other models (Fig ? or Sup Table ? that give quantitative info on this). Figure \ref{fig:acc} shows the decoded state probabilities for both the fine- and coarse- scales for one of the simulation run.

%%% dive duration %%%
For the emission distributions of dive duration, Figure \ref{fig:MLE_dist} shows the empirical joint density of $\hat \mu$ and $\hat \sigma$ for the CarHHMM-DFT.

@@

%%% Acceleration %%%
For acceleration ($Z^{*(1)}$), both the CarHMM-DFT and the CarHHMM-DFT regularly converged to the correct parameters with very little standard error. However, the estimated standard error based on the observed Fisher information regularly overestimated the empirical standard error of $\hat \mu$ for both of these models. 
%This likely is due to the very large auto-correlation terms ($\phi^{(1)} = 0.99$ and $\phi^{(2)} = 0.95$) since as $\phi \to 1$, the value of $\mu$ does not matter in the likelihood evaluation as $\phi \to 1$ implies that $\mathbb{E}(Z^{*(1)}_{t,s^*}) = \phi*(Z^{*(1)}_{t,s^*-1}) + (1-\phi)\mu \to Z^{*(1)}_{t,s^*-1}$. 
The HHMM-DFT regularly overestimates the variance $Z^{*(1)}$ since it does not incorporate auto-correlation into the emission distribution of $Z^{*(1)}$. The CarHHMM has large biases in many of its parameter estimates, especially in sub-dive state 2. 
%because it sees the ``wigliness" of the acceleration data as raw acceleration rather than as Fourier modes, which deflates the value of $\hat \phi$ and inflates the value of $\hat \sigma$.

The supplementary material provides a detailed breakdown of parameter values for acceleration emission distributions and the empirical joint densities for all models and features.

@@

\fi
